D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-08 ootavacuous.tex
Rename Appendix A and forward-reference it
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add section on tooling
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@

All of these tools can incur high overhead, especially manual code
inspection.
However, as we will see, C++ implementations need not use these tools.
Their current analysis is guanteed to be sufficient.
Furthermore, as we will also see, in some special but commonly occurring
circumstances, much simpler methods exist that incur negligible overhead.

% @@@ We could follow this long-established tradition of proposing
% solutions to the OOTA problem without a definition of this problem.
% However, we (perhaps foolishly) felt the need to provide a clear
% definition,

% We therefore define an OOTA cycle (also known as a causal cycle) as a
% cycle in the memory-model sense involving only reads-from links, that
% is, the value stored in on thread being read by the next thread.
% Within each thread in the cycle, the value stored depends on
% the value loaded.
% This dependency might be a control dependency
% (see Listing~\ref{lst:OOTA Cycle})
% or it might be a data dependency
% (see Listing~\ref{lst:Causality Test Case 4} on
% page~\pageref{lst:Causality Test Case 4}).\footnote{
%	Sufficiently motivated readers should have no problem constructing
%	examples involving address dependencies.}
% @@@ There must be at least two distinct values such that if a given load
% returns a given value, that value forces the cycle to execute so as to
% cause the corresponding store's value to be equal to that given value, and
% for that store to occur early enough for the load to return that value.

@@

The following sections discuss these constraints, starting with the
laws of physics, continuing with hardware constraints, continuing
further with constraints imposed by the standard, and culminating
with constraints imposed by the C++ implementations themselves.

@@

Carrying out this type of optimization requires that the implementation
classify dependencies as semantic or not.
Given the limited resources available, almost all implementations will
choose to approximate sdep, which entails some misclassifications, with
the consequences summarized in
Table~\ref{tab:Semantic-Dependency Classification}.
The first row of the table depicts the case where the implementation
correctly identifies a semantic dependency, and thus exhibits temporal
behavior for the corresponding code.
The second row depicts the case where the implementation misclassifies
the dependency as non-semantic, which will result in an incorrect
execution.
In this case, the proper course of action is to fix that bug.
The third row depicts the case where the implementation misclassifies
the dependency as semantic, thus missing a chance to optimize.
The resulting execution is nevertheless correct, and is also temporal.
The fourth and final row depicts the case where the implementation
correctly classifies the dependency as non-semantic, in which case
the execution is both correct and atemporal.

@@

Therefore, in the limiting case, analyses of OOTA cycles that focus on a
given implementation can simply use that implementation's classification
of dependencies as semantic on the one hand or non-semantic on the other.
While taking this approach sacrifices significant generality, it has the
benefit of greatly reduces the cost of dependency analysis, potentially
to zero.

@@

@@

This section has shown that an implementation that correctly
handles the memory model, observable behavior, and the as-if rule
cannot form OOTA cycles.

@@

\section{Demonstration of OOTA-Cycle Definition}
\label{app:Demonstration of OOTA-Cycle Definition}

@@

% @@@ TODO:
% Add a "Tools" section to the real-world constraints section.
%	Note the inversion from compiler, and the ability to complain
%	when analysis cannot be carried out accurately.
% Update wording.
% Add more detail on blocking UB to "Aside on Undefined Behavior".

>>>>

Other OOTA-related litmus tests may be found in
Appendix~\ref{app:OOTA-Related Litmus Tests}.

@@

All of these tools can incur high overhead, especially manual code
inspection.
However, as we will see, C++ implementations need not use these tools.
Their current analysis is guaranteed to be sufficient, courtesy of the
real-world constraints discussed in the next section.
Furthermore, as we will also see, in some special but commonly occurring
circumstances, much simpler methods exist that incur negligible overhead.

@@

The following sections discuss these constraints, starting with the
laws of physics, continuing with hardware constraints, continuing
further with constraints imposed by the standard, and culminating
with constraints imposed by the C++ implementations themselves.
An additional section discusses the OOTA-cycle implications for
any tooling that exists separately for these C++ implementations.

@@

Carrying out this type of optimization requires that the implementation
classify dependencies as semantic or not.
Given the limited resources available, almost all implementations will
choose to approximate sdep, which entails some misclassifications, with
the consequences summarized in
Table~\ref{tab:Semantic-Dependency Classification}.
The first row of the table depicts the case where the implementation
correctly identifies a semantic dependency, and thus exhibits temporal
behavior for the execution at hand of the corresponding code.
The second row depicts the case where the implementation misclassifies
the dependency as non-semantic, which will result in an incorrect
execution.
In this case, the proper course of action is to fix that bug.
The third row depicts the case where the implementation misclassifies
the dependency as semantic, thus missing a chance to optimize.
The resulting execution is nevertheless correct, and is also temporal.
The fourth and final row depicts the case where the implementation
correctly classifies the dependency as non-semantic, in which case
the execution is both correct and atemporal.

@@

@@

\subsection{Semantic Dependencies and Tooling}
\label{sec:Semantic Dependencies and Tooling}

Where a C++ implementation's primary function is to correctly execute
a C++ program, the function of tooling is often to evaluate properties
of possible executions of that same program.
This difference in function results in different sensitivities to
sdep misclassifications, as shown in
Table~\ref{tab:Semantic-Dependency Classification For Tools}.

\begin{table}
\centering
\begin{tabular}{c|c|l}
Actual Code	& Tooling		& Result \\
\hline
sdep		& sdep			& Temporal \\
\cline{2-3}
		& $\neg$sdep		& False Positive (Warning?) \\
\hline
$\neg$sdep	& sdep			& False Negative (Warning?) \\
\cline{2-3}
		& $\neg$sdep		& Atemporal \\
\end{tabular}
\caption{Semantic-Dependency Classification For Tools}
\label{tab:Semantic-Dependency Classification For Tools}
\end{table}

Any misclassification will result in either a false positive or
a false negative, except that many tools give some indication of a
misclassification, for example, printing a warning message indicating
that the code is too complex for it to analyze.
Some projects would choose to restructure the code in such cases on the
grounds that code that is difficult for an automated tool to understand
is likely also being misunderstood by its all-too-human developers.

Some other tools respond to complexity by consuming all available CPU
and memory, which can also be interpreted as a good and sufficient
warning message.

However, there is an important special case for tools that are closely
associated with a C++ implementation.
Such tools can simply use that implementation's classification of
dependencies as semantic on the one hand or non-semantic on the other.
While taking this approach sacrifices significant generality, it has
the not insignificant benefit of greatly reducing the cost of dependency
analysis, potentially all the way to zero.

@@

This section has shown that an implementation that correctly
handles the memory model, observable behavior, and the as-if rule
cannot form OOTA cycles when running on real systems in this
universe.

@@

\section{OOTA-Related Litmus Tests}
\label{app:OOTA-Related Litmus Tests}

@@

% @@@ TODO:
% Update wording.
% Add more detail on blocking UB to "Aside on Undefined Behavior".
